---
title: "Lab 1b. Species Distribution Modeling - Logistic Regression"
author: "Alex Vand"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, messages = FALSE)
```

## 1. Explore

```{r}
librarian::shelf(
  DT, dplyr, dismo, GGally, here, readr, tidyr)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = F)

dir_data    <- here("data/sdm")
```


```{r}
pts_env_csv <- file.path(dir_data, "pts_env.csv")

pts_env <- read_csv(pts_env_csv)
nrow(pts_env)
```

```{r}
datatable(pts_env, rownames = F)
```

## 10) Plot of ggpairs

```{r}
GGally::ggpairs(
  select(pts_env, -ID),
  aes(color = factor(present)))
```

## 2. Logistic Regression

### 2.1 Setup Data

```{r}
# setup model data
d <- pts_env %>% 
  select(-ID) %>%  # remove terms we don't want to model
  tidyr::drop_na() # drop the rows with NA values
nrow(d)
```

## 11) Linear model output with range of values

```{r}
# fit a linear model
mdl <- lm(present ~ ., data = d)
summary(mdl)
```

```{r}
y_predict <- predict(mdl, d, type="response")
y_true    <- d$present

range(y_predict)
```

```{r}
range(y_true)
```

## 12) Generalized Linear Model (GLM)

```{r}
# fit a generalized linear model with a binomial logit link function
mdl <- glm(present ~ ., family = binomial(link="logit"), data = d)
summary(mdl)
```

```{r}
y_predict <- predict(mdl, d, type="response")

range(y_predict)
```

## 13) GLM termp plots

```{r}
# show term plots
termplot(mdl, partial.resid = TRUE, se = TRUE, main = F, ylim="free")
```


## 14) Generalized Additive Model (GAM)

```{r}
librarian::shelf(mgcv)

# fit a generalized additive model with smooth predictors
mdl <- mgcv::gam(
  formula = present ~ s(WC_alt) + s(WC_bio1) + 
    s(WC_bio2) + s(ER_tri) + s(ER_topoWet) + s(lon) + s(lat), 
  family = binomial, data = d)
summary(mdl)
```

## 15) GAM term plots

```{r}
# show term plots
plot(mdl, scale=0)
```

## 16) Question:
*Which GAM environmental variables, and even range of values, seem to contribute most towards presence (above 0 response) versus absence (below 0 response)?*


## 17) Maxent (maximum entropy)

```{r}
# load extra packages
librarian::shelf(
  maptools, sf)

mdl_maxent_rds <- file.path(dir_data, "mdl_maxent.rds")

# show version of maxent
if (!interactive())
  maxent()
```


```{r}
# get environmental rasters
# NOTE: the first part of Lab 1. SDM - Explore got updated to write this clipped environmental raster stack
env_stack_grd <- file.path(dir_data, "env_stack.grd")
env_stack <- stack(env_stack_grd)
plot(env_stack, nc=2)
```

## 18) Maxent variable contribution plot

```{r}
# get presence-only observation points (maxent extracts raster values for you)
obs_geo <- file.path(dir_data, "obs.geojson")
obs_sp <- read_sf(obs_geo) %>% 
  sf::as_Spatial() # maxent prefers sp::SpatialPoints over newer sf::sf class

# fit a maximum entropy model
if (!file.exists(mdl_maxent_rds)){
  mdl <- maxent(env_stack, obs_sp)
  readr::write_rds(mdl, mdl_maxent_rds)
}
mdl <- read_rds(mdl_maxent_rds)

# plot variable contributions per predictor
plot(mdl)
```

## 19) Maxent term plots

```{r}
# plot term plots
response(mdl)
```


```{r}
# predict
y_predict <- predict(env_stack, mdl) #, ext=ext, progress='')

plot(y_predict, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

## 20) Question:
*Which Maxent environmental variables, and even range of values, seem to contribute most towards presence (closer to 1 response) and how might this differ from the GAM results?*
